\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}

% copied from riscv-isa-manual
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}

% Commands for register format figures.

% New column types to use in tabular environment for instruction formats.
% Allocate 0.18in per bit.
\newcolumntype{I}{>{\centering\arraybackslash}p{0.18in}}
% Two-bit centered column.
\newcolumntype{W}{>{\centering\arraybackslash}p{0.36in}}
% Three-bit centered column.
\newcolumntype{F}{>{\centering\arraybackslash}p{0.54in}}
% Four-bit centered column.
\newcolumntype{Y}{>{\centering\arraybackslash}p{0.72in}}
% Five-bit centered column.
\newcolumntype{R}{>{\centering\arraybackslash}p{0.9in}}
% Six-bit centered column.
\newcolumntype{S}{>{\centering\arraybackslash}p{1.08in}}
% Seven-bit centered column.
\newcolumntype{O}{>{\centering\arraybackslash}p{1.26in}}
% Eight-bit centered column.
\newcolumntype{E}{>{\centering\arraybackslash}p{1.44in}}
% Ten-bit centered column.
\newcolumntype{T}{>{\centering\arraybackslash}p{1.8in}}
% Twelve-bit centered column.
\newcolumntype{N}{>{\centering\arraybackslash}p{2.02in}}
% Fifteen-bit centered column.
\newcolumntype{M}{>{\centering\arraybackslash}p{2.2in}}
% Sixteen-bit centered column.
\newcolumntype{K}{>{\centering\arraybackslash}p{2.88in}}
% Twenty-bit centered column.
\newcolumntype{U}{>{\centering\arraybackslash}p{3.6in}}
% Twenty-bit centered column.
\newcolumntype{L}{>{\centering\arraybackslash}p{3.6in}}
% Twenty-five-bit centered column.
\newcolumntype{J}{>{\centering\arraybackslash}p{4.5in}}


\newcommand{\instbit}[1]{\mbox{\scriptsize #1}}
\newcommand{\instbitrange}[2]{~\instbit{#1} \hfill \instbit{#2}~}

\newcommand{\wiri}{\textbf{WIRI}}
\newcommand{\wpri}{\textbf{WPRI}}
\newcommand{\wlrl}{\textbf{WLRL}}
\newcommand{\warl}{\textbf{WARL}}

% end of copied from riscv-isa-manual

\newcommand{\note}{\vspace{1em}\textbf{NOTE: }}
\newcommand{\noteend}{\vspace{1em}}
\newcommand{\consider}{\textbf{CONSIDER: }}

\newcommand{\idWidth}{{6}}
\newcommand{\iommucap}{\textit{iommucap}}
\newcommand{\iommucapen}{\textit{iommucapen}}
\newcommand{\rsidlen}{\textit{RSIDLEN}}
\newcommand{\rsiddiv}{\textit{rsiddiv}}
\newcommand{\ersiddiv}{\textit{ersiddiv}}
\newcommand{\dtbase}{\textit{dtbase}}
\newcommand{\iommucause}{\textit{iommcause}}
\newcommand{\ftval}{\textit{ftval}}
\newcommand{\ftdevid}{\textit{ftdevid}}
\newcommand{\invltlb}{\textit{invltlb}}
\newcommand{\iommucfg}{\textit{iommucfg}}
\newcommand{\resume}{\textit{resume}}
\newcommand{\resumetype}{\textit{resumetype}}
\newcommand{\iommuintno}{\textit{iommuintno}}
\newcommand{\iommuinten}{\textit{iommuinten}}
\newcommand{\dtlen}{\textit{DTLEN}}
\newcommand{\dte}{\textit{dte}}


\begin{document}

\title{Xuantie IOMMU Specification Version 1.0 (Draft)}
\author{Siqi Zhao}
\maketitle


This document describes an IOMMU design specification for the Xuantie platform, which is
based on harts of the RISC-V architecture.

An IOMMU is used to translate virtual addresses in DMA requests from external devices
connected on certain peripheral bus to physical addresses. Abstractly, for all addresses
that can be accessed by an entity identified by a source ID, the IOMMU maintains a relation
from the tuple (source ID, virtual address) to a physical address.

The source ID identifies the entity that initiates DMA operations. The exact meaning of
the entity is dependent on the peripheral bus that the IOMMU is
interfaced with. For example, when the IOMMU is used to translate virtual addresses from
a PCI-Express device, the entity may refers to the \textit{device functions} in the
PCI-Express device. For simplicity, the rest of this document refers to such an entity as
a device. It is also assumed that one device is only capable of generating DMA operations
tagged with one unique identifier.

\note It is up to the implementation to define the exact entity that the term 'device'
refers to. \noteend

The IOMMU utilizes a table structure to maintain the aforementioned
mapping relation. The table structure could be implemented as register arrays or as memory
resident data structure. The table structure consists of multiple levels of tables. There are two
types of tables, the upper levels are IOMMU-specific look-up tables indexed by the source
IDs. The lower level tables contain entries that are the same as the page table entries as
defined in the RISC-V privileged architecture.

\note Typically, peripheral devices assigned to a virtual machine can also send interrupts
that target the harts of that virtual machine, and these interrupts need to be routed as
well, however, this design only specifies the DMA address translations. For interrupts,
the Advanced Interrupt Architecture (AIA) already includes the capability to redirect
interrupts from the devices to specific harts.\noteend

The address translation can be performed with either one stage or two stages. An operating
system kernel may utilize one-stage translation to directly assigned a device to a
process. A hypervisor may utilize the second stage of the two-stage translation to assign
a device to a virtual machine (VM), while the kernel in the VM may utilize the first stage
to further assign the device to processes. Any stage of the translation can be identity
mapping, i.e., no translation in effect.

The IOMMU can optionally implement a translation buffer that caches recent translation
entries. 

\include{arch}
\include{interface}
\include{dev_config}
\include{addr_trans}
\include{addr_tables}
\include{fault_reporting}
\include{tlb}
\include{ordering}
\include{iopmp}
\include{pcie}
\include{list_of_registers}

\end{document}
